Global patterns of forest autotrophic carbon fluxes

### 2020-06-13 manuscript/manuscript draft.Rmd
worked on HM comments

The effects of latitude and climate on FACF were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The basic model for all analyses included a fixed effect of latitude or climate and a random effect of plot nested within geographic area. Geographic areas--*i.e.*, spatially clustered sites--are defined within ForC using a hierarchical cluster analysis on the distance matrix of the sites and a cutoff of 25km [@anderson-teixeira_forc_2018]. We experimented with inclusion of altitude as a fixed effect, but excluded it from the final models because it added very little explanatory power--that is, the difference in AIC ($\Delta$AIC) relative to models excluding altitude was generally small (often $\Delta$AIC<2). Hypotheses were accepted if the $\Delta$AIC between a model including the fixed effect of interest and a corresponding null model excluding that fixed effect exceeded 2.0. All $R^2$ values presented here are marginal $R^2$ values, and refer to the proportion of variation explained by only the fixed effects. Specific analyses are as described below.

We first examined the relationship between latitude and FACF (Q1; Table 1). We tested models with latitude as a linear term (corresponding null: model without latitude) and as a second-order polynomial term (corresponding null: model with latitude as a linear term), and calculated AIC values to determine the best model. **BBL: careful about terminology here. In statistics a "linear model" includes polynomial models, which are linear in their parameters. Thus I (and many readers perhaps) find this confusing.** Models were accepted as the best model if $\Delta$AIC > 2 with respect to the corresponding null, and were significant with respect to a null model with no fixed term.

To test whether trends in component fluxes across latitude sum to match those of larger fluxes, regression lines for smaller component fluxes were summed to generate new estimates of larger fluxes, which were then compared against the latitudinal regression of the larger flux. Confidence intervals for the larger flux were calculated using the 'bootMer' function from the lme4 package [@bates_fitting_2015]. This analysis was applied to the following sets of fluxes: (1) $GPP = NPP + R_{auto}$, (2) $NPP = ANPP + BNPP$, and (3) $ANPP = ANPP_{foliage} + ANPP_{stem}$. In addition, we estimated total belowground C flux (TBCF, not analyzed due to limited data) as $TBCF = BNPP + R_{root}$.

Variation in allocation to component carbon fluxes along latitudinal gradients was explored for the following pairings: $GPP:NPP$, $ANPP:BNPP$, $ANPP_{foliage}:ANPP_{stem}$, $ANPP_{foliage}:NPP$, $ANPP_{stem}:NPP$, $ANPP:NPP$, and $BNPP:NPP$. For each set of paired fluxes, measurements taken at the same site and plot, and in the same year, were paired together, and the ratio of each pair of measurements calculated. The ratios were regressed against latitude and climate variables, using the linear model specified above. Cook's distance analyses were carried out for each of the models, and indicated that data from a few high-elevation sites were having a disproportionate influence on the regressions. To account for this, models were re-run using only data from sites $\le$ 1000m.

@@

Multivariate models were used to investigate the potential joint and interactive effects of climate variables on carbon fluxes. We explored a large range of climate variable combinations and ultimately decided to focus on MAT and MAP. We favored these over variables with similar explanatory power because they are the most commonly reported climate variables. MAT was consistently among the best individual predictors of FACF (Table S2), although it was occasionally out-performed by closely correlated climate variables (Fig. S2). In combination with MAT as a primary climate variable, MAP was consistently among the best secondary climate variables. Specifically, compared against a model including MAT x MAP, there were no other variables that, in interaction with MAT, improved AIC by >2.0 and $R^2$ by >0.05. In examining the joint and interactive effects of climate variables, an additive model including MAT and MAP as fixed effects was accepted when $\Delta$AIC >2 relative to a null including only MAT as a fixed effect. An interactive model including an MAT x MAP interaction was accepted when $\Delta$AIC >2 relative to a null including MAT and MAP as fixed effects.

To test whether and how FACF varied with climate when standardised by growing season length (Q5), we first standardized FACF by dividing by growing season length (as defined above). We then tested for correlations between these standardised fluxes and growing season climate variables, using only linear models.

@@

In total, we analyzed 1228 records from nine FACF variables taken from forests that had experienced no major anthropogenic disturbances within the past 100 years. These records represented a total of 154 distinct geographic areas across all forested biogeographic and climate zones (Fig. 1, Table 2).

@@

All major carbon fluxes decreased linearly with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes--particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). The FACF that were most poorly predicted by latitude were $BNPP_{fine.root}$ ($R^2$=0.17) and $ANPP_{stem}$ ($R^2$=0.18). For all FACF, the relationship with latitude was best predicted by the first-order linear model.

```{r echo=FALSE, out.width='100%', fig.cap = "Latitudinal trends in forest autotropic carbon flux. Lines of best fit are plotted according to the best model selected during analysis. All regressions are significant $(p<0.05)$. Each panel shows major C fluxes together with component fluxes. Also plotted are predicted trends in the major C fluxes based on the sum of component fluxes. 95\\% confidence intervals are plotted for the major flux for comparison with predicted trends. In (d),  which shows three belowground fluxes, the major flux, total belowground carbon flux, is one for which we have no data", fig.pos='H'}

@@

There was little evidence of substantial variation in C allocation across latitude or climate (Fig. S3). Of the 7 FACF ratios regressed against latitude and three climate variables (MAT, MAP, temperature seasonality), there were only five signficant relationships, all with $R^2 \le 0.4$ (Fig. S3). Specifically, the proportion of NPP allocated to $ANPP_{foliage}$ decreased with latitude ($R^2=0.32$) and increased with MAT ($R^2=0.37$), and the proportion of $NPP$ allocated aboveground ($ANPP$) decreased weakly with latitude ($R^2=0.11$) and temperature seasonality ($R^2=0.17$), while increasing with MAT ($R^2=0.11$).  The remaining 23 relationships examined were non-significant.

>>>>

The effects of latitude and climate on FACF were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The basic model for all analyses included a fixed effect of latitude or climate and a random effect of plot nested within geographic area. Geographic areas--*i.e.*, spatially clustered sites--are defined within ForC using a hierarchical cluster analysis on the distance matrix of the sites and a cutoff of 25km [@anderson-teixeira_forc_2018]. We experimented with inclusion of altitude as a fixed effect, but excluded it from the final models because it added very little explanatory power--that is, the difference in AIC ($\Delta AIC$) relative to models excluding altitude was generally small (often $\Delta AIC$<2). Effects were considered signficant when inclusion of the term of interest resulted in $\Delta AIC \ge 2.0$ relative to a corresponding null model. All $R^2$ values presented here are marginal $R^2$ values, and refer to the proportion of variation explained by only the fixed effects. Specific analyses are as described below.

We first examined the relationship between latitude and FACF (Q1; Table 1). We tested models with latitude as a first-order linear term (corresponding null: model without latitude) and as a second-order polynomial or logarithmic term (corresponding null: model with latitude as a first-order linear term). We selected as the best model that with the highest $\Delta AIC$ relative to a null model with no fixed term.

To test whether trends in component fluxes across latitude sum to match those of larger fluxes, regression lines for smaller component fluxes were summed to generate new estimates of larger fluxes. 
We then determined whether these summed predictions fell within the **95%** CI for the larger flux across the entire range included latitude. **(Becky, is 95% the correct number)**
Confidence intervals for the larger flux were calculated using the 'bootMer' function from the lme4 package [@bates_fitting_2015]. **(Helene asks for more details here: what does the function do? are CIs just based on uncertainty in regression?)**
This analysis was applied to the following sets of fluxes: (1) $GPP = NPP + R_{auto}$, (2) $NPP = ANPP + BNPP$, and (3) $ANPP = ANPP_{foliage} + ANPP_{stem}$. In addition, we estimated total belowground C flux (TBCF, not analyzed due to limited data) as $TBCF = BNPP + R_{root}$.

Variation in allocation to component carbon fluxes along latitudinal gradients was explored for the following pairings: $NPP:GPP$, $ANPP:BNPP$, $ANPP_{foliage}:ANPP_{stem}$, $ANPP_{foliage}:NPP$, $ANPP_{stem}:NPP$, $ANPP:NPP$, and $BNPP:NPP$. **[issue 91](https://github.com/forc-db/Global_Productivity/issues/91)**
For each set of paired fluxes, measurements taken at the same site and plot, and in the same year, were paired together, and the ratio of each pair of measurements calculated. The ratios were regressed against latitude and climate variables, using the linear model specified above. 
Cook's distance analyses were carried out for each of the models, and indicated that data from a few high-elevation sites were having a disproportionate influence on the regressions. Thus, models were re-run using only data from sites $\le$ 1000m.

@@

Multivariate models were used to investigate the potential joint and interactive effects of MAT and MAP on carbon fluxes. An additive model including MAP in addition to MAT was accepted when $\Delta AIC$ >2 relative to a null including only MAT as a fixed effect. An interactive model including an MAT x MAP interaction was accepted when $\Delta$AIC >2 relative to a null including MAT and MAP as fixed effects.

To test whether and how FACF varied with climate when standardised by growing season length (Q5), we first standardized FACF by dividing by growing season length (as defined above). We then tested for correlations between these standardised fluxes and growing season climate variables, using only first-order linear models.

@@

In total, we analyzed 1228 records from nine FACF variables taken from forests that had experienced no major anthropogenic disturbances within the past 100 years. These records represented a total of **#** plots in 154 distinct geographic areas across all forested biogeographic and climate zones (Fig. 1, Table 2).

@@

All major carbon fluxes decreased linearly with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes--particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). The FACF that were most poorly predicted by latitude were $BNPP_{fine.root}$ ($R^2$=0.17) and $ANPP_{stem}$ ($R^2$=0.18). For all FACF, the relationship with latitude was better fit by the first-order linear model than by second-order polynomial or logarithmic models.

```{r echo=FALSE, out.width='100%', fig.cap = "Latitudinal trends in forest autotropic carbon flux. Lines of best fit are plotted according to the best model selected during analysis. All regressions are significant $(p<0.05)$. Each panel shows major C fluxes together with component fluxes. Also plotted are predicted trends in the major C fluxes based on the sum of component fluxes. 95\\% confidence intervals are plotted for the major flux for comparison with predicted trends. In (d), which shows three belowground fluxes, the major flux, total belowground carbon flux, has insufficient data (n=9) to support a regression", fig.pos='H'}

@@

There was little evidence of systematic variation in C allocation with latitude or climate (Fig. S3). Of 28 relationships tested (7 FACF ratios regressed against latitude, MAT, MAP, and temperature seasonality), there were only five signficant relationships, all with $R^2 \le 0.4$ (Fig. S3). The proportion of NPP allocated to $ANPP_{foliage}$ decreased with latitude ($R^2=0.32$) and increased with MAT ($R^2=0.37$). The proportion of $NPP$ allocated aboveground ($ANPP$) decreased weakly with latitude ($R^2=0.11$) and temperature seasonality ($R^2=0.17$), while increasing with MAT ($R^2=0.11$).
